Structure comparison for multivariate data by\ldots / with application to \ldots

### 2017-02-27 article/article_v0.01.tex
Suggestion for new title + correction of typos

\title{Something that is not PCADSC}

@@

\emph{Principal component analysis} (PCA) may be interpreted in terms of the \emph{rank-q-reconstruction error}. To describe this consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, and let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ be their average. Suppose that we want to describe each of the observations by $q$ numbers instead of the original $d$ numbers. The rank-q-reconstruction error is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*} 
PCA ensures the existence of the subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ of $S$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a sign whenever the eigenvalues are different. If some of the eigenvalue are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{i=q+1}^d \lambda_i$. In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called the \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called the \emph{scores}. The $j$th loading can also found as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon: u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2
\end{align*}

>>>>

%\title{Something that is not PCADSC}
\title{Structure comparison for multivariate data by\ldots / with application to \ldots}

@@

\emph{Principal component analysis} (PCA) may be interpreted in terms of the \emph{rank-q-reconstruction error}. To describe this consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, and let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ be their average. Suppose that we want to describe each of the observations by $q$ numbers instead of the original $d$ numbers. The rank-q-reconstruction error is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*} 
PCA ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ of $S$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a sign whenever the eigenvalues are different. If some of the eigenvalue are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{i=q+1}^d \lambda_i$. In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}. The $j$th loading can also found as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2
\end{align*}
